---
title: "Gapminder"
author: "Your Awesome Data Science Professor"
format: html
editor: visual
---

## Setup

This code chunk sets up the analysis by loading the key libraries.

```{r setup}
library(tidyverse)
library(gapminder)
```

## Look at the data

Let's view the `gapminder` dataset. (This is interactive, so un-comment the `View` command.)

```{r}
View(gapminder)
head(gapminder)
```

## Starter graph

Let's start with a plot of GDP per capita vs. life expectancy in Asia in 2007. (This is based off the example in the slides.)

```{r}
gapminder |> 
  filter(continent == "Asia", year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + 
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy", 
       title = "Life expectancy versus GDP in Asia in 2007") +
  theme_bw()
```

## Your Turn 1a: GDP vs Life Expectancy in 1957 in Europe

I've copy-and-pasted the above code here. Modify this code to instead plot GDP per capita versus life expectancy in 1957 in Europe.

```{r}
gapminder |> 
  filter(continent == "Europe", year == 1957) |> 
  ggplot(aes(x = gdpPercap, y = lifeExp))+
  geom_point()+
  scale_x_log10()+
  labs( x = "GDP Per Capita", y = "Life Expectancy",
        title = "Life Expectancy Versus GDP Per Capita in Europe in 1957")+
  theme_bw()

```

## Your Turn 1b: GDP vs Life Expectancy in 1957 in the Americas

I've again copy-and-pasted the above code here. Modify this code to instead plot GDP per capita versus life expectancy in 1957 in the Americas. (*Hint:* Look at the data to learn the right value for `continent`.)

```{r}
gapminder |> 
  filter(continent == "Americas", year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + 
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy", 
       title = "Life expectancy versus GDP in Asia in 2007") +
  theme_bw()
```

## Your Turn 2a: Life expectancy over time in the US

I've again copy-and-pasted the above code here. Modify this code to plot life expectancy over time in the United States. *Hint 1:* Plot `year` on the x-axis and life expectancy (`lifeExp`) on the y-axis. *Hint 2:* What do you need to change under `filter`?

```{r}
gapminder |> 
  filter(continent == "Asia", year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + 
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy", 
       title = "Life expectancy versus GDP in Asia in 2007") +
  theme_bw()
```

## Your Turn 2b: Life expectancy over time in the US (cont'd)

Wouldn't that plot look better as a line rather than points? Take a stab at converting the plot from points to a line by modifying `geom_point` .

```{r}
gapminder |> 
  filter(continent == "Americas") |> 
  ggplot(aes(x = year, y = lifeExp))+
  geom_smooth()+
  scale_x_log10()+
  labs (x = "Year", "Life Expectancy",
        title = "Life Expectancy Over Time in the US")+
  theme_bw()
```
